---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(tibble)
```

# R/`nomatch`

> A matching alternative for more efficiently evaluating effectiveness of interventions using observational data

---

## Description

The `nomatch` package uses a G-computation style estimator to compute the effectiveness of a binary intervention in observational cohort studies that use the target trial emulation approach. The proposed estimator tends to produce similar point estimates as matching-based estimators but can be more efficient.

---

## Installation

You can install the development version of `nomatch` with:

``` r
# install.packages("devtools")
devtools::install_github("ewu16/nomatch")
```

---

## Example

This minimal example shows how to use `nomatch` to obtain cumulative incidences and 
their derived effect measures such as risk differences (RD), risk ratios (RR), and relative risk reductions (1 - RR).

We use a simple simulated dataset based on an observational vaccine study,
although data from other disease settings can be used. 

```{r example, fig.width = 8, fig.height = 4}

# Load package
library(nomatch)

# View example data
simdata <- as_tibble(simdata) #for prettier printing 
head(simdata)
```

The dataset contains the following:

-  one row per individual (`ID`)
-  a set of baseline covariates (`x1`, `x2`)
-  exposure status (`V`) with values `1 = vaccinated, 0 = unvaccinated`. 
-  time to vaccination (`D_obs`); for unvaccinated individuals this is left as `NA`. 
- right censored survival data `(Y, event)` 
   - `Y` represents follow-up time for an outcome such as infection, hospitalization or death.
   - `event` indicates whether individual experienced the event or censoring with values `1 = event, 0 = censored`.


```{r, message = FALSE, warning = FALSE}

# Use nomatch to compute cumulative incidence and effect estimates 
fit <- nomatch(data = simdata,
               outcome_time = "Y",
               outcome_status = "event",
               exposure = "V",
               exposure_time = "D_obs", 
               covariates = c("x1", "x2"),
               immune_lag = 14,
               timepoints = seq(30, 180, by = 30),
               boot_reps = 10)

# Examine object produced by nomatch - shows risk ratio estimates by default 
fit        

# Plot cumulative incidence and effect estimates
plot(fit) 
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an issue](https://github.com/ewu16/nomatch/issues).

---

## Citation

If you use the `nomatch` package in your work, please cite the following:

```{r, include = FALSE}
c <- citation("nomatch")
toBibtex(c)
```

```
 @Manual{,
   title = {nomatch: Estimate Effectiveness of Interventions in Target Trial Emulation Observational Studies
             Without Using Matching},
   author = {Emily Wu},
   year = {2026},
   https://github.com/ewu16/nomatch},
   url = {https://ewu16.github.io/nomatch/},
 }
 ```
 

