---
title: "nomatch"
output: rmarkdown::html_vignette
bibliography: ../inst/references.bib
vignette: >
  %\VignetteIndexEntry{nomatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3
)
```


```{r core_pkgs, echo=FALSE, warning = FALSE, message = FALSE}
library(tibble)
library(ggplot2)
library(dplyr)
```

## Introduction

The `nomatch` package uses G-computation style estimation to compute
marginal cumulative incidence in observational cohort studies with a binary exposure and a time-to-event outcome. 
The approach aligns with the target trial emulation framework [@hernan2016_using] and provides
an alternative to matching-based methods, which can be inefficient.
Estimation involves fitting two exposure-specific Cox models which adjust for baseline confounders.
Predictions from these models are used to estimate marginal 
cumulative incidence under interventions of exposure and no exposure.
Effect measures such as risk differences, risk ratios, and relative risk reduction are also provided. 

---

## Method 

To study the effectiveness of an intervention, we would ideally conduct a randomized trial. In such a trial, individuals are assigned to receive the intervention or not at a well-defined time, and outcomes are measured from the time of assignment. Cumulative incidence between the two groups can then be compared at a fixed follow-up time.

In observational studies, this comparison is less straightforward because individuals who remain unexposed do not have an obvious start of follow-up. One way to address this issue is to imagine assigning individuals to receive (or not receive) the intervention on specific days, similar to the assignment time in a randomized trial.

Under this perspective, we can define cumulative incidences for each intervention day and covariate group. These quantities can then be averaged (marginalized) over the observed distribution of exposure days and covariates to obtain overall cumulative incidence estimates. This mirrors the marginal comparisons typically reported in randomized trials, where cumulative incidence estimates implicitly average over the distribution of assignment times and baseline covariates within each treatment group.

---

## Usage

We use `nomatch` to estimate the effectiveness of a
binary exposure in an observational cohort study. The package can be loaded as follows: 

```{r setup}
library(nomatch)
```

We illustrate the use of the package using a simple simulated dataset, `simdata`, 
included in the package. For exposition, we will assume this dataset represents data from an 
observational vaccine study. In practice, data from other disease settings can be used as well. 
The first few rows of `simdata` can be viewed as follows:


```{r}
simdata <- as_tibble(simdata) #for prettier printing

# View data
head(simdata)
```


The data contains one row per individual (`ID`) and a set of individual baseline covariates (`x1`, `x2`).
Information on an individual's exposure is encompassed in  1) their binary vaccination status `V` with values `1 = vaccinated, 0 = unvaccinated`, 
and 2) their time to receiving vaccination `D_obs`, which is a numeric value for vaccinated individuals and 
`NA` for unvaccinated individuals.
The data also includes a right-censored survival outcome`(Y, event)` where `Y` represents follow-up time for an outcome (e.g. infection or death), and `event` indicates whether the individual experienced the event with values `1 = event, 0 = censored`. 

<!-- We will assume that time to vaccination `D_obs` and time to outcome `Y` are measured relative -->
<!-- to a calendar date of study start because calendar time is -->
<!-- often an important time scale in vaccine studies due to changing infection dynamics. In other settings, the relevant time scale may be -->
<!-- time since enrollment or other meaningful event, or age. Regardless of the choice of time scale,  -->
<!-- the same time origin should be used to define the time to exposure and time to outcome variables,  -->
<!-- as well as the baseline covariates.  -->

---

### Estimating cumulative incidence

We can invoke the `nomatch` function to compute marginal cumulative incidences. 
The main types of arguments are:

- **Data arguments**`- data, outcome_time, outcome_status, exposure, exposure_time, covariates`: Specify the dataset and the variable names describing the outcome, exposure, and confounders. 

- **Estimation arguments** control what cumulative incidences are computed: 
  - `immune_lag` excludes events occuring with `immune_lag` time units after exposure. This is common in vaccine studies, where immunity takes time to develop. In this example we set `immune_lag = 14` days, although it can be set to 0 when not relevant. 
  - `timepoints` when to evaluate cumulative incidence. Here we compute cumulative incidence at 30, 60, 90, ..., 180 days after vaccination.

- **Bootstrap argument** - `boot_reps`controls how many bootstrap replicates are used to construct confidence intervals and p-values. 
We use a small number here for speed, but `boot_reps = 1000` is recommended for publication-quality results. By default, Wald-type bootstrap confidence intervals are returned.


```{r}

# Compute cumulative incidence 
fit <- nomatch(data = simdata,
               outcome_time = "Y",
               outcome_status = "event",
               exposure = "V",
               exposure_time = "D_obs", 
               covariates = c("x1", "x2"),
               immune_lag = 14,
               timepoints = seq(30, 180, by = 30),
               boot_reps = 10)


```

The cumulative incidence estimates (`cuminc_0` for unexposed and  `cuminc_1` for exposed) ) and other effectiveness measures
are stored within the `$estimates` component of the fitted object. 

```{r}
str(fit$estimates, give.attr = FALSE)
```


We can thus preview the relative risk reduction estimates (i.e. vaccine effectiveness) using 

```{r}
head(fit$estimates$relative_risk_reduction) 
```


If all confounders are adjusted for, the resulting cumulative incidence/effect estimates can be interpreted as the cumulative incidences or effect that would be observed in a clinical trial in which participants 1) are similar to those observed to be exposed and 2) who enroll into the study such that the distribution of enrollments times matches the distribution of observed exposure times. 

### Additional details about estimation approach 

Internally, `nomatch()` fits two Cox models — one for the unexposed, one for the exposed.

In the **unexposed model**, time to endpoint is measured from the study start (or chosen time origin). This model includes all individuals, with exposed individuals censored at their time of exposure. The model adjusts fo  baseline covariates.

In the **exposed model**, time to endpoint is measured from the time of exposure. This model includes only exposed individuals who remained event-free `immune_lag` days after exposure. The model adjusts for baseline covariates and flexibly accounts for when exposure occurred (by default, using a natural cubic spline with 4 degrees of freedom).


These Cox models are used to predict time- and covariate-specific cumulative incidences which are then marginalized.
By default, the predictions are marginalized over the distribution of exposure times and covariates in the exposed.


A summary of the estimation approach and fitted models can be obtained by 
```{r}
summary(fit)
```


and the weights used for marginalization are returned in the `$weights` component of the fitted object. 

### Plotting cumulative incidence curves

We can visualize the cumulative incidence and effectiveness estimates using `plot()`. 
If smooth cumulative incidence curves are desired, it is best to supply a dense grid of timepoints. 
Our example plot appears somewhat coarse because we evaluated the estimates at only six timepoints.

```{r}
plot(fit, effect = "relative_risk_reduction")
```

#### Plotting simultaneous confidence intervals 


The plot above shows pointwise confidence intervals, which are the usual 95% CIs computed independently at each timepoint. 
If inference on the entire curve is desired, simultaneous confidence intervals should be used instead.
These can be added using `add_simultaneous_ci()`. This function updates an existing fit with simultaneous confidence intervals, provided bootstrap samples were retained (keep_boot_samples = TRUE, the default). 
The updated fit should be saved as a new object. 


```{r}
#Compute simultaneous CI
fit_with_simul <- add_simultaneous_ci(fit, seed = 1234)
```

The simultaneous confidence intervals are now stored in the fitted object. 
```{r}
fit_with_simul
```


To plot simultaneous confidence intervals, use `plot()` with the argument `ci_type = "simul"`.  
The fitted object must already contain the simultaneous confidence intervals; otherwise, the call will return an error.

```{r}
#Plot simultaneous confidence bands 
plot(fit_with_simul , effect = "relative_risk_reduction", ci_type = "simul") 
```

The simultaneous confidence intervals indicate that 95% of such bands, constructed across repeated samples, would contain the entire true curve.

### Creating custom plots

Users may also wish to create their own plots. The function `estimates_to_df()`
takes the estimates component of the fitted object and reformats it into a long dataset suitable for plotting.

```{r}
plot_data <- estimates_to_df(fit)

head(as_tibble(plot_data))
```

For example, we can create a plot that overlays the cumulative incidence estimates
for the two exposure types as follows:

```{r}
plot_data |> 
  filter(term %in% c("cuminc_0", "cuminc_1")) |> 
  ggplot(aes(x = t0, y = estimate, color = term, fill = term)) + 
  geom_line() + 
  geom_ribbon(aes(ymin = wald_lower, ymax = wald_upper), alpha = 0.2, linewidth = 0) + 
  theme_bw() +
  scale_color_discrete(breaks = c("cuminc_0", "cuminc_1"), labels = c("Unvaccinated", "Vaccinated"))+
  scale_fill_discrete(breaks = c("cuminc_0", "cuminc_1"), labels = c("Unvaccinated", "Vaccinated"))+ 
  labs(x = "Time since vaccination",
       y = "Estimate",
       title = "Cumulative Incidence Curves", 
       color = "Exposure Group",
       fill = "Exposure Group")
  
```



### Comparison with matching 

Our estimation approach is generally expected to produce
similar point estimates as a rolling-cohort matching analysis  in which exposed individuals are matched to unexposed individuals at their time of exposure. However, our approach may offer greater precision, which can be particularly beneficial in smaller samples or when endpoints are rare.

For comparison, we provide a simple implementation of a rolling-cohort matching analysis using 1:1 exact matching and Kaplan–Meier estimation to compute marginal cumulative incidences. These matching functions are intentionally limited in scope and may be slow on large datasets.


```{r}
# ------------------------------------------------------------------------------
# 3. Compare results with matching estimator

matched_cohort <- match_rolling_cohort(data = simdata,
                                       outcome_time = "Y",
                                       exposure = "V",
                                       exposure_time = "D_obs", 
                                       matching_vars = c("x1", "x2"),
                                       id_name = "ID",
                                       seed = 5678)

matched_data <- matched_cohort[[1]]


fit_matching <-matching(matched_data = matched_data,
                        outcome_time = "Y",
                        outcome_status = "event",
                        exposure = "V",
                        exposure_time = "D_obs", 
                        immune_lag = 14,
                        timepoints = seq(30, 180, by = 30),
                        boot_reps = 10) 

fit_matching

# Plot matching vs proposed estimator - nomatch tends to have similar point estimates but narrower
# confidence intervals

comparison  <- rbind(estimates_to_df(fit) |> mutate(method = "nomatch (G-computation)"),
                       estimates_to_df(fit_matching)  |> mutate(method = "matching"))

comparison |> 
  filter(term %in% c("cuminc_0", "cuminc_1", "relative_risk_reduction")) |> 
  mutate(term_label = factor(term, 
                             c("cuminc_0", "cuminc_1", "relative_risk_reduction"),
                             c("Cumulative Incidence:\n Unvaccinated",
                               "Cumulative Incidence:\n Vaccinated",
                               "Relative Risk Reduction"))) |> 
  ggplot(aes(x = t0, y = estimate, color = method, fill = method)) + 
  geom_line() +
  geom_ribbon(aes(ymin = wald_lower, ymax = wald_upper), alpha = 0.2, linewidth = 0.1) + 
  facet_wrap(~term_label, scales = "free") + 
  ggh4x::facetted_pos_scales(
    y = list(
      ggplot2::scale_y_continuous(labels = scales::label_percent(),
                                  limits = c(0, 0.11)),
      ggplot2::scale_y_continuous(labels = scales::label_percent(),
                                  limits = c(0, 0.11)),
      ggplot2::scale_y_continuous(labels = scales::label_percent(),
                                  limits = c(0, 0.8))
    )) + 
  theme_bw() +
  theme(strip.background = ggplot2::element_rect(fill = "white", color = "white"),
         strip.text = ggplot2::element_text(size = 11, colour = "black")) + 
  labs(x = "Time since vaccination", 
       y = "Estimate", 
       color = "Method",
       fill = "Method")
```





## Session Information

```{r seesionInfo, echo = FALSE}
sessionInfo()
```




